### 2013-11-21 manuscript/manuscript/ms_main.tex
Initial commit

Arid rangelands are at high risk of sudden desertification. This process of persistent vegetation loss under livestock grazing is hardly understood but evidently an issue of the system's bistability. Under particular environmental conditions and grazing regimes, two ecosystem states are likewise resilient to perturbations: One state with a robust vegetation cover of shrubs and grasses as well as a desert with no significant vegetation cover (Scheffer 2001, Aguiar and Sala 1999). After crossing critical thresholds, or tipping-points, the system transites between the two stable states.

For arid ecosystems, the major mechanism behind this bistability was found to be local facilitation of plant individuals. The presence of a plant provides its direct local environment with abilities of water retention, organic matter accumulation and protection against external stressors. As a consequence, the vegetation grows in patches (Aguiar and Sala 1999).  When implemented in a simple spatially explicit model, this positive neighborhood effect provides true bistability on the landscape scale (K\'efi et al 2007a,b).

This parent model was used before to investigate the consequences of increased plant mortality. This simple formulation of additional plant stress can be interpreted as the effects of increased grazing intensity by large herbivores (K\'efi et al 2007a). However, the model is lacking some important features of grazing: the additional individual plant mortality (1) did not vary with vegetation density, and (2) it applied globally, affecting all plants likewise. 
Both assumptions are justified by the need for model simplicity, but disallow a translation of the model parameters to real livestock rates. The relationship between livestock rates and the stability of vegetation cover is a crucial information for management decisions. The present study aims at increasing the realism and applicability of the model by applying two extensions that enable simulations of a grazing gradient on drylands.

%Grazing depends on vegetation cover
The first extension adds a contrasting assumption on the plant density dependence of feeding to the parent model. The prior model defined grazing as a constant risk of additional plant death. On the population level, grazing mortality increased with vegetation cover (K\'efi et al. 2007). However, grazers forage in the landscape and consume to fulfill their daily demands. In other words, the number of plants suffering from grazing is rather constant with vegetation cover. As a consequence, at low vegetation cover the grazing risk on individual plants will be high, whereas when vegetation cover is high, risk will be shared among many plants.
In reality, the total feeding of large herbivores in the landscape is not entirely independent of vegetation cover but follows a non-linear saturating curve (type-II functional response, Holling 1959, Spalinger and Hobbs 1992). Thus, particularly at low vegetation cover grazing is limited by the search success of the grazer, which is resource-density dependent. In rangeland systems, however, the functional response at low vegetation cover depends on factors on a larger scale: the animals move between landscapes motivated by internal decision patterns or by external management decisions (Senft 1989). We assume therefore, that smooth type-II functional responses, as described for wild ungulates, are unlikely to happen in intensive rangelands within the spatial scale adressed in the present study.

Because of these uncertainties and for reasons of simplicity we contrast the parent assumption of density independent individual risk with density independent global feeding. We expect constant global feeding to consolidate both stable states: the desert state will be more resilient against recolonisation of single plants while the vegetated state will provide better protection to the plant individual. As a consequence, the reversibility of catastrophic shifts is expected to deteriorate.

%Spalinger and Hobbs 1992

%Grazing is spatialy differentiated
The second extension implements a local differentiation of grazing pressure. %For some time, the relationship between vegetation structure and grazing is acknowledged to be interdependent. Grazing affects the mortality of plants as well as plant recruitment, soil suitability, and other factors in positive as well as negative ways. Vegetation cover is a 
%Senft et al 1987
%Bailey 1996
Particularly, the importance of direct neighborhood interactions on the individual plant scale (local scale) in explaining spatial structure was demonstrated by Adler et al. (2001). The authors showed in a theoretical framework similar to the present study, that positive local interactions are superior to environmental heterogeneity or heterogeneous stress in explaining the patterns of plant spatial heterogeneity (Adler et al 2001). 
To differentiate the grazing locally, we implement an associative protection of plant individuals growing next to each other. This mechanism is observed frequently among shrubs in dryland ecosystems and considered to be a major driver of spatial heterogeneity of stress on local scales.

 %Bisigato et al 2005 Ecography
%Milchunas and Noy-Meir 2002

Here, the individual vulnerability against grazing is highest for plants growing isolated. The more direct neighbors a plant associates to, the lower is the effect of grazing. A plant in the patch center is not affected by grazing. We expect this mechanism to benefit the desert state, because at low densities only few plants are associated.

% comparison and main objectives/questions

 We provide a full factorial comparison of the consequences of these alternative models on vegetation structure and the ecosystems' bistability properties.

\section{Methods}
We used a cellular automaton to investigate the consequences of different grazing assumptions on the structural properties of the vegetation. This model was used before (Kefi et al 2007a, b) and can be resolved numerically in spatially explicit simulations. It defines landscape as a grid of cells, each can potentially take one of three discrete cell states: 'vegetated' cells are occupied by a plant individual (annotated as '+' in equations; black cells in figures); 'enriched' cells are suitable for seeds to germinate and colonise the cell, but do not contain adult plants right now  ('0' , grey cells); 'degraded' cells represent bare ground with lacking organic matter and bad water retention, and therefore can not be colonised by arriving seeds  ('$-$', white cells).
Transitions of cells states are only possible between vegetated and enriched (plant death / recolonisation) as well as enriched and degraded (degradation / regeneration). In biological terms, a degraded spot needs to be enriched first, before a plant can grow. Vice versa, when a plant dies, it leaves the spot in an enriched state, which might become degraded later. The rates for these transitions are defined in the following paragraphs. They might be constant values or functions of the global or local plant vegetation density (= $\rho_+$, global vegetation cover; $n_+$) .

@@

We applied these rules on a grid of 100 $\times$ 100 cells of 0.25m$^2$. Each timestep is defined as one year. The applicable transition probabilities for each single cell given in equations 1, 2, 3 and 4 are compared against uniform random numbers between 0 and 1 to determine if a transition occurs or if the cell remains unchanged. To initialize the grid with a standardized patchy vegetation structure, each model simulation was run for 500 years without any grazing (g = 0) using an initial environmental value $b_ini \in \left\{ 0.25, 3, 0.8, 0.9 \right\}$ before grazing is taking action. %(Note: I want to complete this to a highly resoluted gradient, to investigate the unstable equilibria). 
Starting from the preconditioned grids, the four grazing models (Table 1) with varying grazing intensity ($g \in \left{ 0, 0.025, 0.05, 0.075 , 0.10 \right}$) were run over a gradient of environmental quality ($b \in \left{ 0.2, ... , 1.0 \right}$) for 1.500 timesteps.

An average vegetation cover, $\rho_+$, as well as the average cell mortality,$\hat{m}$ , is determined as the average density of all grids of the final 500 timesteps.

@@

\begin{figure}[h]%
\includegraphics[width=\columnwidth]{figures/stability.pdf}%
\caption{Stable states of the vegetation cover, $\rho_+$, over gradient of environmental quality, $b$, under different grazing models. a) parent model with static, global grazing mortality, $g_0/0.5$. b) livestock model with density dependend, global grazing mortality, $g_0 / \rho_+$. c) associative protection model with differentiated grazing at patch border, $g_0/0.5 \frac{v}{\hat{v}}$ . d) livestock \& associative protection model with density dependend, differentiated grazing at patch border,  $g_0/ \rho_+ \frac{v}{\hat{v}}$ . Gradient of grazing intensity from grey: $g_0 = 0$, black to red: $g_0 \in \left\{ 0.025, 0.05, 0.075, 0.1 \right\}$.}%
\label{}%
\end{figure}

\begin{figure}[h]%
\includegraphics[width=\columnwidth]{figures/patterns.pdf}%
\caption{Stable states of the vegetation cover, $\rho_+$, over gradient of environmental quality, $b$, under different grazing models. a) parent model with static, global grazing mortality, $g_0/0.5$. b) livestock model with density dependend, global grazing mortality, $g_0 / \rho_+$. c) associative protection model with differentiated grazing at patch border, $g_0/0.5 \frac{v}{\hat{v}}$ . d) livestock \& associative protection model with density dependend, differentiated grazing at patch border,  $g_0/ \rho_+ \frac{v}{\hat{v}}$ . Gradient of grazing intensity from grey: $g_0 = 0$, black to red: $g_0 \in \left\{ 0.025, 0.05, 0.075, 0.1 \right\}$.}%
\label{}%
\end{figure}

@@

In reality, however, the functional response of large grazers on vegetation, particularly at low vegetation cover, is dynamic. In rangelands, management decisions and spatial responses of the grazers will reduce feeding pressure at very low densities.

In fact, the local effects of large herbivores on plants are manifold and qualitatively different. While direct feeding reduces biomass in the first place, it might lead to growth enhancement and increased productivity. On the local scale, vegetation benefit from the presence of herbivores through fertilization, inoculation with seeds, 
In our model, however, we simplify these factors to a linear negative impact of grazer presence on the local vegetation. The

@@

\section{APPENDIX}

\subsection{simulations}

The grid was defined to have periodic borders, connecting the east border to the west, the north to the south.
Prerun for 100 years. 
during pre-run, environmental conditions $b_ini \in \left\{ 0.250, 0.274, 0.300, 0.329, 0.360, 0.395, 0.433, 0.474, 0.520, 0.570, 0.624, 0.684, 0.749, 0.821, 0.900 \right\} $ emerge a patch structure ranging from low to high vegetation cover. Trajectories towards a stable vegetated state or towards desert are documented and used to approximate the unstable equilibrium.

, the whole grid is updated synchronously multiple times, to increase stochasticity
each time step is devided by the number of updates per year ( = 5) and

\subsection{model fitting}
 Four models were compared on each snapshot of t

\begin{table}[t!]
\label{tab:fitmodels}
\caption{ Models describing patch size distribution }
\centering
\begin{tabular}{lll}

\toprule
 & function & log-transform \\ \cmidrule(rl){2-2} \cmidrule(rl){3-3}
Power Law &  $n = a s^{- \alpha}$ & $ \log{n} =\log{a} - \alpha \log s$\\
Limited Power Law &  $n = b + a s^{- \alpha} $ & $ \log{n} = \log a - \alpha \log{s} + \log(1+\frac{b}{a s^{-\alpha} })$\\
Truncated Power Law & $n = a s^{- \alpha} \mathrm{e}^{\frac{s}{S_x}}$  & $\log{n} = \log a  - \alpha  \log s  - \frac{s}{Sx}  $ \\
Exponential &  $n = a \mathrm{e}^{-\varepsilon s} $ & $ \log{n} =  \log{a} - \varepsilon s $\\
	\bottomrule
\multicolumn{3}{p{12.5cm}}{\footnotesize  }
	\end{tabular}
\end{table}

table with log-transformed models

>>>>

Drylands are at high risk of sudden desertification. This process of persistent vegetation loss is hardly understood but evidently a consequence of bistability. Under particular environmental conditions, two ecosystem states are likewise resilient to perturbations: One state with high vegetation cover of shrubs and grasses as well as a desert with no significant vegetation cover (Scheffer 2001). After crossing critical thresholds, or tipping-points, the system transites between the two stable states.

For the dryland ecosystem, the major mechanism behind this bistability was found to be local facilitation of plant individuals. The presence of a plant provides its direct local environment with abilities of water retention, organic matter accumulation and protection against external stresses. When implemented in simple spatial models, This positive neighborhood effect provides true bistability on the landscape scale (K\'efi et al 2007a,b).

This parent model was used before to investigate the consequences of increased plant mortality. This simple formulation of additional plant stress can be interpreted as the effects of increased grazing intensity by large herbivores (K\'efi et al 2007a). However, it is lacking some important features of grazing: the added plant mortality (1) applied globally, affecting all plants likewise, and (2) did not vary with vegetation density. The present study aims at increasing the realism of the model by applying two extensions that produce a more realistic gradient of grazing pressure on drylands. We provide a full factorial comparison of the consequences of these alternative models on vegetation structure and the ecosystems' bistability properties. The first extension adds the assumption of density dependent feeding. Grazers forage in the landscape and consume to fulfill their demands. At low density the grazing risk on individual plants will be high, while it will be shared among many plants when vegetation cover is high. For reasons of simplicity, we contrast the prior assumption of density independent individual risk with density independent global feeding. We expect this to consolidate the desert state by making it more resilient against recolonisation of single plants. The second extension implements an associative protection of plant individuals growing next to each other, a mechanism observed frequently among shrubs in dryland ecosystems. Here, the individual vulnerability against grazing is highest for plants growing isolated. The more direct neighbors a plant associates to, the lower is the effect of grazing. A plant in the patch center is not affected by grazing. We expect this mechanism to benefit the desert state, because at low densities only few plants are associated.

\section{Methods}
We used a cellular automaton to investigate the consequences of different grazing assumptions on the structural properties of the vegetation. This model was used before (Kefi et al 2007a, b) and can be resolved numerically in spatially explicit simulations but also analytically in differential equations. It defines landscape as a grid of cells, each can potentially take one of three discrete cell states: 'vegetated' cells are occupied by a plant individual (annotated as '+' in equations; black cells in figures); 'enriched' cells are suitable for seeds to germinate and colonise the cell, but do not contain adult plants right now  ('0' , grey cells); 'degraded' cells represent bare ground with lacking organic matter and bad water retention, and therefore can not be colonised by arriving seeds  ('$-$', white cells).
Transitions of cells states are only possible between vegetated and enriched (plant death / recolonisation) as well as enriched and degraded (degradation / regeneration). In biological terms, a degraded spot needs to be enriched first, before a plant can grow. Vice versa, when a plant dies, it leaves the spot in an enriched state, which might become degraded later. The rates for these transitions are defined in the following paragraphs. They might be constant values or functions of the global or local plant vegetation density (= $\rho_+$, global vegetation cover; $n_+$) .

@@

We apply these rules on a grid of 100 $\times$  100 cells of 0.25m$^2$ . The grid is defined to have periodic borders, connecting the east border to the west, the north to the south. Each timestep, which is defined as one year, the whole grid is updated synchronously multiple times, to increase stochasticity. At each update, the applicable transition probabilities for each single cell given in equations 1, 2, 3 and 4 are devided by the number of updates per year ( = 5) and tested against a uniform random number between 0 and 1. To initialize the grid with a standardized patch structure, each model simulation runs for 1.000 timesteps without any grazing using an initial environmental value $b_ini \in \left\{ 0.25, 3, 0.8, 0.9 \right\}$ before grazing is taking action (Note: I want to complete this to a highly resoluted gradient, to investigate the unstable equilibria). On those preconditioned grids, the four grazing models (Table ) run for 1.500 timesteps. An average vegetation cover, $\rho_+$, as well as the average cell mortality,$\hat{m}$ , is determined from the grids of the final 100 timesteps.

@@

\begin{figure}%
\includegraphics[width=\columnwidth]{figures/stability.pdf}%
\caption{Stable states of the vegetation cover, $\rho_+$, over gradient of environmental quality, $b$, under different grazing models. a) parent model with static, global grazing mortality, $g_0/0.5$. b) livestock model with density dependend, global grazing mortality, $g_0 / \rho_+$. c) associative protection model with differentiated grazing at patch border, $g_0/0.5 \frac{v}{\hat{v}}$ . d) livestock \& associative protection model with density dependend, differentiated grazing at patch border,  $g_0/ \rho_+ \frac{v}{\hat{v}}$ . Gradient of grazing intensity from grey: $g_0 = 0$, black to red: $g_0 \in \left\{ 0.025, 0.05, 0.075, 0.1 \right\}$.}%
\label{}%
\end{figure}

@@

@@
